---
title: "Borough Demographics Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: flatly
---

By borough / zip code (like the TA’s example) 
Primary use of subway 
Average subway ride length 
Most used line 

Perception of affordability 
Opinion on increased fair 
Satisfaction 

Converting zip codes into boroughs. I used https://www.nyc.gov/assets/planning/download/office/data-maps/nyc-population/census2000/sf1p11.xls to get majority of the borough's zip codes. Any remaining ones, I looked up on Google. There were three zip codes that I could not find a match for, and were likely miswritten, so I removed those data rows. 


```{r setup, include=TRUE}
library(tidyverse)
library(plotly)
library(dplyr)
```


```{r}
#Clean data
subway_rider_data = 
  read_csv("data/2019_subway_rider_data.csv") |> 
  janitor::clean_names()


subway_drops =
  subway_rider_data |> 
  select(-survey_stop_borough, -survey_stop_location) |> 
  drop_na() |> 
  mutate(survey_id = row_number()) |> 
  relocate(survey_id, .before = 1)
  
rider_data =
  subway_drops |>
  mutate(
    use_of_subway_frequency = case_when(
      use_of_subway_frequency == "Everyday / Almost everyday" ~ "Every day",
      use_of_subway_frequency == "More than once a week, varies" ~ "Several times a week",
      use_of_subway_frequency %in% c("Weekdays, Mon - Fri", "~5 days per week") ~ "5 days per week",
      use_of_subway_frequency == "Once a week" ~ "Once a week",
      use_of_subway_frequency == "Once a month" ~ "Once a month",
      use_of_subway_frequency == "Almost never" ~ "Almost never",
      use_of_subway_frequency == "Never" ~ "Never",
      TRUE ~ NA_character_
    ),
    use_of_subway_frequency = 
      factor(
        use_of_subway_frequency,
        levels = c("Never", 
                   "Almost never", 
                   "Once a month", 
                   "Once a week",
                   "Several times a week", 
                   "5 days per week", 
                   "Every day"),
        ordered = TRUE)
    )


rider_data =
  subway_drops |>  
  mutate(average_length_subway_ride = 
           factor(average_length_subway_ride,
                  levels = c("< 20 min", 
                             "20 - 40 min", 
                             "40 - 60 min",
                             "60 - 90 min", 
                             "90 min - 2 hours"),
                  ordered = TRUE))


rider_data =
  subway_drops |>
  mutate(
    overall_satisfaction = 
      recode(
        overall_satisfaction,
        "Not satisfied" = "Unsatisfied"),
    overall_satisfaction = 
      factor(
        overall_satisfaction,
        levels = c("Highly unsatisfied",
                 "Unsatisfied",
                 "Neutral",
                 "Somewhat satisfied",
                 "Very satisfied"),
      ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(
    frequency_of_delays = factor(
      frequency_of_delays,
      levels = c("Everyday",
                 "A few times a week",
                 "Rarely",
                 "Almost never",
                 "Never"),
      ordered = TRUE)
    )
    
#Zip code --> borough 

zip_code_data = 
  read_csv("data/nyc_zipcodes.csv") %>%
  select(borough, zipcode)

#Make sure character type
rider_data <- rider_data %>%
  mutate(zip_code = as.character(zip_code))
zip_code_data <- zip_code_data %>%
  mutate(zipcode = as.character(zipcode))

#Join
rider_data_full <- rider_data %>%
  left_join(zip_code_data, by = c("zip_code" = "zipcode"))

#If there's no match, we want to keep the zip_code in the borough column, so we use coalesce to handle this
rider_data_full <- rider_data_full %>%
  mutate(borough = coalesce(borough, as.character(zip_code)))

#manually go through remaining zip codes - removing 11135 11290 12105 as they are not real zip codes & also removing one observation that didn't have zip code 
rider_data_full <- rider_data_full %>%
  filter(zip_code != 12105) %>%
  filter(zip_code != 11290) %>%
  filter(zip_code != 11135) %>%
  filter(survey_id != 919) 
```


### Chart A 
```{r, fig.height=14}
#Location distribution 
borough_df <- rider_data_full %>% 
  count(borough)

plot_ly(
  borough_df,
  x = ~borough,
  y = ~n,
  type = "bar",
  hovertemplate = "Borough: %{x}<br>Count: %{y}<extra></extra>"
) %>%
  layout(title = "Survey Respondents by Borough",
         xaxis = list(title = "Borough"),
         yaxis = list(title = "Count")) %>%
  layout(autosize = TRUE)

```


### Chart B
```{r, fig.height=14}
#Primary use of subway 
primary_df <- rider_data_full %>%
  count(borough, primary_use_of_subway) %>%
  group_by(borough) %>%
  mutate(prop = n / sum(n))

plot_ly(
  primary_df,
  x = ~borough,
  y = ~prop,
  color = ~primary_use_of_subway,
  type = "bar",
  hovertemplate = 
    "Borough: %{x}<br>Use: %{customdata}<br>Percent: %{y:.1%}<extra></extra>",
  customdata = ~primary_use_of_subway
) %>%
  layout(
    barmode = "stack",
    title = "Primary Subway Use by Borough",
    yaxis = list(tickformat = ".0%", title = "Proportion")
  ) %>%
  layout(autosize = TRUE)
```



### Chart C
```{r, fig.height=14}
#Average subway ride length 
ride_df <- rider_data_full %>%
  count(borough, average_length_subway_ride)

plot_ly(ride_df,
        x = ~average_length_subway_ride,
        y = ~n,
        color = ~borough,
        type = "bar",
        hovertemplate = "Borough: %{color}<br>Ride Length: %{x}<br>Count: %{y}<extra></extra>") %>%
  layout(
    barmode = "stack",
    title = "Ride Length Distribution by Borough",
    xaxis = list(title = "Ride Length Category"),
    yaxis = list(title = "Count")
  )
```


### Chart D
```{r, fig.height=14}
#Most used line 
line_df <- rider_data_full %>%
  count(borough, subway_line_used_most_often)

#Make the hover show counts per borough
hover_df <- line_df %>%
  group_by(subway_line_used_most_often) %>%
  summarise(
    total_count = sum(n),
    borough_info = paste0(borough, ": ", n, collapse = "<br>")
  )

#Produce plot
plot_ly(hover_df) %>%
  add_segments(
    x = 0,
    xend = ~total_count,
    y = ~reorder(subway_line_used_most_often, total_count),
    yend = ~reorder(subway_line_used_most_often, total_count),
    hoverinfo = "none"
  ) %>%
  add_markers(
    x = ~total_count,
    y = ~reorder(subway_line_used_most_often, total_count),
    hovertemplate = "Line: %{y}<br>Total Count: %{x}<br>%{customdata}<extra></extra>",
    customdata = ~borough_info
  ) %>%
  layout(
    title = "Most Used Subway Lines by Borough",
    xaxis = list(title = "Count"),
    yaxis = list(title = "Line")
  )
```


### Chart E
```{r, fig.height=14}
#Perception of affordability 
afford_df <- rider_data_full %>%
  count(borough, is_subway_affordable) %>%
  group_by(borough) %>%
  mutate(prop = n / sum(n))

plot_ly(
  afford_df,
  x = ~borough,
  y = ~prop,
  color = ~is_subway_affordable,
  type = "bar",
  hovertemplate = 
    "Borough: %{x}<br>Affordability: %{customdata}<br>Percent: %{y:.1%}<extra></extra>",
  customdata = ~is_subway_affordable
) %>%
  layout(
    barmode = "stack",
    title = "Affordability Perception by Borough",
    yaxis = list(tickformat = ".0%", title = "Proportion")
  ) %>%
  layout(autosize = TRUE)

```


### Chart F
```{r, fig.height=14}
#Opinion on increased fair 
library(ggplot2)
library(dplyr)
library(patchwork)  # for arranging plots

# Prepare data
satis_df <- rider_data_full %>%
  count(borough, opinion_on_increased_fares) %>%
  group_by(borough) %>%
  mutate(
    prop = n / sum(n)
  ) %>%
  ungroup()

# Unique boroughs
boroughs <- unique(satis_df$borough)

# Create individual pie charts for each borough
plots <- lapply(boroughs, function(b) {
  df_b <- filter(satis_df, borough == b)
  
  ggplot(df_b, aes(x = "", y = prop, fill = opinion_on_increased_fares)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    labs(title = b, fill = "Opinion") +
    theme_void() +
    theme(legend.position = "bottom")
})

# Arrange plots in a grid (e.g., 2 rows)
wrap_plots(plots, nrow = 2)

```

### Chart G
```{r, fig.height=14}
#Satisfaction 
satis_overall <- rider_data_full %>%
  count(overall_satisfaction) %>%
  mutate(prop = n / sum(n),
         percent = paste0(round(prop*100,1), "%"))

plot_ly(
  satis_overall,
  labels = ~overall_satisfaction,
  values = ~prop,
  type = "pie",
  hole = .5,
  textinfo = "label+percent",
  hovertemplate = "%{label}<br>%{percent}<extra></extra>"
) %>%
  layout(title = "Aggregated Satisfaction Levels") %>%
  layout(autosize = TRUE)

```


