---
title: "subway_map"
output: html_document
date: "2025-12-02"
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)
library(lubridate)
library(readxl)
library(gt)
```

Here I will clean the 2019_subway_rider_data 

```{r}
subway_rider_data = 
  read_csv("data/2019_Subway_Rider_Survey_20251121.csv") |> 
  janitor::clean_names()


subway_drops =
  subway_rider_data |> 
  select(-survey_stop_borough, -survey_stop_location) |> 
  drop_na() |> 
  mutate(survey_id = row_number()) |> 
  relocate(survey_id, .before = 1)
  
```

Now I will specifically clean the data within the columns

Cleaned rider frequency table: 
```{r}
rider_data =
  subway_drops |>
  mutate(
    use_of_subway_frequency = case_when(
      use_of_subway_frequency == "Everyday / Almost everyday" ~ "Every day",
      use_of_subway_frequency == "More than once a week, varies" ~ "Several times a week",
      use_of_subway_frequency %in% c("Weekdays, Mon - Fri", "~5 days per week") ~ "5 days per week",
      use_of_subway_frequency == "Once a week" ~ "Once a week",
      use_of_subway_frequency == "Once a month" ~ "Once a month",
      use_of_subway_frequency == "Almost never" ~ "Almost never",
      use_of_subway_frequency == "Never" ~ "Never",
      TRUE ~ NA_character_
    ),
    use_of_subway_frequency = 
      factor(
        use_of_subway_frequency,
        levels = c("Never", 
                   "Almost never", 
                   "Once a month", 
                   "Once a week",
                   "Several times a week", 
                   "5 days per week", 
                   "Every day"),
        ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(average_length_subway_ride = 
           factor(average_length_subway_ride,
                  levels = c("< 20 min", 
                             "20 - 40 min", 
                             "40 - 60 min",
                             "60 - 90 min", 
                             "90 min - 2 hours"),
                  ordered = TRUE))


rider_data =
  subway_drops |>
  mutate(
    overall_satisfaction = 
      recode(
        overall_satisfaction,
        "Not satisfied" = "Unsatisfied"),
    overall_satisfaction = 
      factor(
        overall_satisfaction,
        levels = c("Highly unsatisfied",
                 "Unsatisfied",
                 "Neutral",
                 "Somewhat satisfied",
                 "Very satisfied"),
      ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(
    frequency_of_delays = factor(
      frequency_of_delays,
      levels = c("Everyday",
                 "A few times a week",
                 "Rarely",
                 "Almost never",
                 "Never"),
      ordered = TRUE)
    )
    
```


Making some of the variables as a character and some as a factor!


```{r}
rider_data = 
  rider_data |> 
  mutate(
    zip_code = as.character(zip_code)
    )


rider_data =
  rider_data |> 
  mutate(
    across(c(subway_line_used_most_often,
             get_to_subway_via,
             primary_use_of_subway,
             alternative_transport), as.factor)
  )
```


```{r}
rider_data = 
  rider_data %>%
  mutate(
    top_three_complaints = 
      str_remove_all(top_three_complaints, "\\[|\\]|'")
  )
```


Converting zip codes into boroughs. I used https://www.nyc.gov/assets/planning/download/office/data-maps/nyc-population/census2000/sf1p11.xls to get majority of the borough's zip codes. Any remaining ones, I looked up on Google and cross-checked. There were four zip codes that I could not find a match for, and were likely miswritten, so I removed those data rows. 

```{r}
zip_code_data = 
  read_csv("data/nyc_zipcodes.csv") %>%
  select(borough, zipcode)


#Make sure character variables 
zip_code_data <- zip_code_data %>%
  mutate(zipcode = as.character(zipcode)) #make sure it's a character variable 
rider_data <- rider_data %>%
  mutate(zip_code = as.character(zip_code))

#Join
rider_data_full <- rider_data %>%
  left_join(zip_code_data, by = c("zip_code" = "zipcode"))

#Fill borough where missing
rider_data_full <- rider_data_full %>%
  mutate(borough = coalesce(borough, as.character(zip_code)))

#Remove invalid zip codes
rider_data_full <- rider_data_full %>%
  filter(survey_id != 2410) %>% 
  filter(survey_id != 944) %>% 
  filter(survey_id != 241) %>% 
  filter(survey_id != 919) 
```

final data set is rider_data_full

These next two pieces of code came from composite_score.rmd
Join-ready Subway Line Directory + Join to cleaned dataset
```{r}
subway_directory <- tribble(
~subway_line, ~group_color, ~hex_color,
"1","Red Line","#EE352E",
"2","Red Line","#EE352E",
"3","Red Line","#EE352E",
"4","Green Line","#00933C",
"5","Green Line","#00933C",
"6","Green Line","#00933C",
"7","Purple Line","#B933AD",
"A","Blue Line","#0039A6",
"C","Blue Line","#0039A6",
"E","Blue Line","#0039A6",
"B","Orange Line","#FF6319",
"D","Orange Line","#FF6319",
"F","Orange Line","#FF6319",
"M","Orange Line","#FF6319",
"N","Yellow Line","#FCCC0A",
"Q","Yellow Line","#FCCC0A",
"R","Yellow Line","#FCCC0A",
"W","Yellow Line","#FCCC0A",
"G","Light Green Line","#6CBE45",
"J","Brown Line","#996633",
"Z","Brown Line","#996633",
"L","Gray Line","#A7A9AC",
"S","Shuttle","#808183"
)
```


```{r}
# Categories defining "long delays"
long_categories <-  c(
  "20 - 40 min", "40 - 60 min", "60 - 90 min",
  "90 min - 2 hours", "> 60 min"
)

# Ordinal score for frequency of delays
delay_freq_scores <- c(
  "Never" = 1,
  "Almost never" = 2,
  "Rarely" = 3,
  "A few times a week" = 4,
  "Everyday" = 5
)

# Numeric midpoints for average delay length
delay_length_scores <- c(
  "< 20 min"        = 10,     # midpoint choice
  "20 - 40 min"     = 30,
  "40 - 60 min"     = 50,
  "60 - 90 min"     = 75,
  "90 min - 2 hours" = 105
)

# Satisfaction ordinal scale
satisfaction_scores <- c(
  "Highly unsatisfied" = 1,
  "Unsatisfied"        = 2,
  "Neutral"            = 3,
  "Somewhat satisfied" = 4,
  "Very satisfied"     = 5
)
```

```{r}
scored_data <- rider_data_full |>
  mutate(
    subway_line = subway_line_used_most_often |>
      str_replace_all("[^A-Za-z0-9]", "/") |>
      str_split("/")
  ) |>
  unnest(subway_line) |>
  mutate(
    subway_line = str_trim(subway_line),
    subway_line = str_to_upper(subway_line)
  ) |>
  filter(subway_line %in% subway_directory$subway_line) |>
  left_join(subway_directory, by = "subway_line") |>
  mutate(
    long_delay       = average_length_subway_ride %in% long_categories,
    delay_freq_num   = delay_freq_scores[frequency_of_delays],
    delay_length_num = delay_length_scores[average_length_subway_ride],   # ← FIX
    satisfaction_num = satisfaction_scores[overall_satisfaction]
  )

```

```{r}
family_scores_raw <- scored_data |>
  group_by(group_color) |>
  summarise(
    pct_long_delay   = mean(long_delay, na.rm = TRUE),
    avg_delay_freq   = mean(delay_freq_num, na.rm = TRUE),
    avg_delay_length = mean(delay_length_num, na.rm = TRUE),
    avg_satisfaction = mean(satisfaction_num, na.rm = TRUE)
  ) |>
  ungroup()

```

```{r}
family_scores_norm <- family_scores_raw |>
  mutate(
    z_pct_long_delay   = scale(pct_long_delay)[,1]    * -1,
    z_delay_freq       = scale(avg_delay_freq)[,1]    * -1,
    z_delay_length     = scale(avg_delay_length)[,1]  * -1,
    z_satisfaction     = scale(avg_satisfaction)[,1]
  )

```

```{r}
family_composite <- family_scores_raw |>
  mutate(
    z_pct_long_delay = scale(-pct_long_delay)[,1],
    z_delay_freq     = scale(-avg_delay_freq)[,1],
    z_delay_length   = scale(-avg_delay_length)[,1],
    z_satisfaction   = scale(avg_satisfaction)[,1]
  ) |>

  # Replace z-score NAs with 0 (neutral)
  mutate(
    across(starts_with("z_"), ~replace_na(.x, 0))
  ) |>

  mutate(
    composite_score = z_pct_long_delay +
                      z_delay_freq +
                      z_delay_length +
                      z_satisfaction
  ) |>
  arrange(desc(composite_score)) |>
  mutate(rank = row_number())
```

```{r}
# Step 2: Remove extra columns & make all scores positive
family_composite_clean <- family_composite |>
  mutate(
    composite_score = composite_score - min(composite_score, na.rm = TRUE)
  ) |>
  arrange(desc(composite_score)) |>
  mutate(rank = row_number()) |>
  select(group_color, composite_score, rank)
```



Back to schedules programming:



```{r}
zip_path = "data/nyu_2451_34758.zip"

if (!file.exists("data/routes_nyc_subway_may2016.shp")) 
  {unzip(zip_path, exdir = "data")}

```


```{r}
subway_routes = 
  sf::st_read("data/routes_nyc_subway_may2016.shp")


subway_routes = 
  subway_routes |>
  sf::st_transform(4326)

```


Using our color palette
```{r}
palette_df = 
  tibble::tribble(
    ~route_id, ~color,
    "1",  "#EE352E",
    "2",  "#EE352E",
    "3",  "#EE352E",
    "4",  "#00933C",
    "5",  "#00933C",
    "6",  "#00933C",
    "7",  "#B933AD",
    "A",  "#0039A6",
    "C",  "#0039A6",
    "E",  "#0039A6",
    "B",  "#FF6319",
    "D",  "#FF6319",
    "F",  "#FF6319",
    "M",  "#FF6319",
    "N",  "#FCCC0A",
    "Q",  "#FCCC0A",
    "R",  "#FCCC0A",
    "W",  "#FCCC0A",
    "G",  "#6CBE45",
    "J",  "#996633",
    "Z",  "#996633",
    "L",  "#A7A9AC",
    "S",  "#808183",

    "FS", "#808183",
    "GS", "#808183",
    "H",  "#808183",
    "SI", "#808183"
    )
```



```{r}
subway_routes =
  subway_routes |>
  left_join(palette_df, by = "route_id") |>
  left_join(
    subway_directory |>
      select(subway_line, group_color),
    by = c("route_id" = "subway_line")
  ) |>
  left_join(
    family_composite_clean,
    by = "group_color"
  ) |>
  mutate(
    medal = case_when(
      rank == 1 ~ " 🥇",
      rank == 2 ~ " 🥈",
      rank == 3 ~ " 🥉",
      TRUE      ~ ""
    ),
    route_display = paste0(route_id, " Train")
  )

```


```{r}

nyc_subway_routes =
  leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolylines(
    data   = subway_routes,
    color  = ~color,
    weight = 2,
    popup  = ~paste0(
      "<b>Route: </b>", route_id, "<br>",
      "<b>Name: </b>", route_long, "<br>",
      "<b>Line Family: </b>", group_color, "<br>",
      "<b>Composite Score: </b>", round(composite_score, 2), "<br>",
      "<b>Rank: </b>", rank, medal
    ),
    label = ~paste0(route_display, " ( Rank ", rank, " ", medal, ")"),
    group  = ~route_id,
    highlightOptions = highlightOptions(
        weight = 5,
        opacity = 1,
        bringToFront = TRUE)
    ) |>
  addLayersControl(
    overlayGroups = palette_df$route_id,
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  htmlwidgets::prependContent(
    tags$div(
      style = "font-size:20px; font-weight:bold; padding:8px;",
      "NYC Subway Routes")
  )

nyc_subway_routes
```







