---
title: "Regression Analysis"
---

Running regression to develop our subway line ranking system. overall_satisfaction = subway_line_used_most_often + approximate_delay_duration + most_common_reason_for_delay + primary_use_of_subway + frequency_of_delays

All predictors are converted to an ordinal, categorical variable. Assumption, overall satisfaction as a Likert scale 1-5 continuous variable, allowing for linear regression. 

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(dplyr)
library(viridis)
library(ordinal)
library(sjPlot)
```


```{r}
#DATA CLEANING
subway_rider_data = 
  read_csv("data/2019_subway_rider_data.csv") |> 
  janitor::clean_names()


subway_drops =
  subway_rider_data |> 
  select(-survey_stop_borough, -survey_stop_location) |> 
  drop_na() |> 
  mutate(survey_id = row_number()) |> 
  relocate(survey_id, .before = 1)
  
rider_data =
  subway_drops |>
  mutate(
    use_of_subway_frequency = case_when(
      use_of_subway_frequency == "Everyday / Almost everyday" ~ "Every day",
      use_of_subway_frequency == "More than once a week, varies" ~ "Several times a week",
      use_of_subway_frequency %in% c("Weekdays, Mon - Fri", "~5 days per week") ~ "5 days per week",
      use_of_subway_frequency == "Once a week" ~ "Once a week",
      use_of_subway_frequency == "Once a month" ~ "Once a month",
      use_of_subway_frequency == "Almost never" ~ "Almost never",
      use_of_subway_frequency == "Never" ~ "Never",
      TRUE ~ NA_character_
    ),
    use_of_subway_frequency = 
      factor(
        use_of_subway_frequency,
        levels = c("Never", 
                   "Almost never", 
                   "Once a month", 
                   "Once a week",
                   "Several times a week", 
                   "5 days per week", 
                   "Every day"),
        ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(average_length_subway_ride = 
           factor(average_length_subway_ride,
                  levels = c("< 20 min", 
                             "20 - 40 min", 
                             "40 - 60 min",
                             "60 - 90 min", 
                             "90 min - 2 hours"),
                  ordered = TRUE))


rider_data =
  subway_drops |>
  mutate(
    overall_satisfaction = 
      recode(
        overall_satisfaction,
        "Not satisfied" = "Unsatisfied"),
    overall_satisfaction = 
      factor(
        overall_satisfaction,
        levels = c("Highly unsatisfied",
                 "Unsatisfied",
                 "Neutral",
                 "Somewhat satisfied",
                 "Very satisfied"),
      ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(
    frequency_of_delays = factor(
      frequency_of_delays,
      levels = c("Everyday",
                 "A few times a week",
                 "Rarely",
                 "Almost never",
                 "Never"),
      ordered = TRUE)
    )
    
#Zip code --> borough 

zip_code_data = 
  read_csv("data/nyc_zipcodes.csv") %>%
  select(borough, zipcode)

#Make sure character type
rider_data <- rider_data %>%
  mutate(zip_code = as.character(zip_code))
zip_code_data <- zip_code_data %>%
  mutate(zipcode = as.character(zipcode))

#Join
rider_data_full <- rider_data %>%
  left_join(zip_code_data, by = c("zip_code" = "zipcode"))

#If there's no match, we want to keep the zip_code in the borough column, so we use coalesce to handle this
rider_data_full <- rider_data_full %>%
  mutate(borough = coalesce(borough, as.character(zip_code)))

#manually go through remaining zip codes - removing 11135 11290 12105 as they are not real zip codes & also removing one observation that didn't have zip code 
rider_data_full <- rider_data_full %>%
  filter(survey_id != 2410) %>% 
  filter(survey_id != 944) %>% 
  filter(survey_id != 241) %>% 
  filter(survey_id != 919) 
```


```{r}
#Converting categorical variables to orders
regression_rider_data <-
  rider_data_full %>%
    mutate(
    average_length_subway_ride = 
      recode(
        average_length_subway_ride,
        "< 20 min" = "0", 
        "20 - 40 min" = "1",
        "40 - 60 min" = "2",
        "60 - 90 min" = "3",
        "90 min - 2 hours" = "4"), 
    average_length_subway_ride = 
      factor(
        average_length_subway_ride, 
        levels = c("0", "1", "2", "3", "4"), 
        ordered = TRUE
      )) %>%
    mutate(
    frequency_of_delays = 
      recode(
        frequency_of_delays,
        "Never" = "0", 
        "Almost Never" = "1",
        "Rarely" = "2",
        "A few times a week" = "3",
        "Everyday" = "4"), 
    frequency_of_delays = 
      factor(
        frequency_of_delays, 
        levels = c("0", "1", "2", "3", "4"), 
        ordered = TRUE
      )) %>%
    mutate(
    approximate_delay_duration = 
      recode(
        approximate_delay_duration,
        "5 - 10 min" = "0", 
        "10 - 20 min" = "1",
        "20 - 45 min" = "2",
        "45 - 60 min" = "3",
        "> 60 min" = "4"), 
    approximate_delay_duration = 
      factor(
        approximate_delay_duration, 
        levels = c("0", "1", "2", "3", "4"), 
        ordered = TRUE
      )) %>%
    mutate(
    frequency_of_rerouting = 
      recode(
        frequency_of_rerouting,
        "Almost never" = "0",
        "Only on weekends" = "1", 
        "Only at night" = "2",
        "All the time" = "3"), 
    frequency_of_rerouting = 
      factor(
        frequency_of_rerouting, 
        levels = c("0", "1", "2", "3"), 
        ordered = TRUE
      )) %>%
     mutate(
      borough = case_when(
      borough == "Manhattan" ~ 0,
      borough %in% c("Brooklyn", "Queens", "Staten Island", "Bronx", "Long Island") ~ 1,
      borough %in% c("New Jersey", "New York State", "Outside NY") ~ 2,
      TRUE ~ NA_real_
    )) 

#Converting overall satisfaction to be "continuous" 
regression_rider_data <-
  regression_rider_data %>%
    mutate(
      overall_satisfaction = case_when(
      overall_satisfaction == "Highly unsatisfied" ~ 1,
      overall_satisfaction == "Not satisfied" ~ 2,
      overall_satisfaction == "Neutral" ~ 3,
      overall_satisfaction == "Somewhat satisfied" ~ 4,
      overall_satisfaction == "Very satisfied" ~ 5))
```



```{r}
#Run the model 
fit <- lm(overall_satisfaction ~ average_length_subway_ride + frequency_of_delays + approximate_delay_duration + frequency_of_rerouting + borough, data = regression_rider_data)

#Tidying output
fit |> 
  broom::tidy() |> 
  select(term, estimate, p.value)

#Visualizations
regression_rider_data |> 
  modelr::add_residuals(fit) |> 
  ggplot(aes(x = overall_satisfaction, y = resid)) + geom_point()
```


We should do a logistic regression model. 

```{r}
#Recode overall satisfaction (outcome) to be ordered 
regression_rider_data <- regression_rider_data %>%
  mutate(
    overall_satisfaction = factor(
      overall_satisfaction,
      levels = c(1, 2, 3, 4, 5),
      ordered = TRUE
    )
  )

#Run logistic regression model 
model <- clm(
  overall_satisfaction ~ 
    average_length_subway_ride +
    frequency_of_delays +
    approximate_delay_duration +
    frequency_of_rerouting +
    borough,
  data = regression_rider_data,
  link = "logit"
)

#Summary of model
model_df <- broom::tidy(model, conf.int = TRUE) |>
  dplyr::mutate(across(where(is.numeric), round, 3)) 

#Publication style table 
tab_model(model)
```

Visualizing effect size - if predictors increase or decrease satisfaction odds.
```{r}
tidy_model <- broom::tidy(model) %>%
  filter(!grepl("\\|", term)) %>%
  mutate(odds_ratio = exp(estimate),
         conf.low = exp(estimate - 1.96 * std.error),
         conf.high = exp(estimate + 1.96 * std.error))

plot_ly(tidy_model,
        x = ~odds_ratio, 
        y = ~term,
        type = 'scatter',
        mode = 'markers',
        error_x = list(type = 'data', symmetric = FALSE,
                       array = tidy_model$conf.high - tidy_model$odds_ratio,
                       arrayminus = tidy_model$odds_ratio - tidy_model$conf.low)) %>%
  layout(title = "Odds Ratios with 95% CI",
         xaxis = list(title = "Odds Ratio"),
         yaxis = list(title = "Predictor"))

```


Distributions  
```{r}
#Pick blue from gradient 
viridis_blue <- viridis(100)[40]

plot_ly(regression_rider_data, 
        x = ~overall_satisfaction, 
        type = "histogram",
         marker = list(color = viridis_blue))
```


Predicted probabilities with average subway length and satisfaction 
```{r}
#Create grid of predicted probabilities 
grid <- expand.grid(
  average_length_subway_ride = levels(regression_rider_data$average_length_subway_ride),
  borough = unique(regression_rider_data$borough),
  frequency_of_delays = levels(regression_rider_data$frequency_of_delays)[1],  # fix at reference level
  approximate_delay_duration = levels(regression_rider_data$approximate_delay_duration)[1],
  frequency_of_rerouting = levels(regression_rider_data$frequency_of_rerouting)[1]
)

grid$average_length_subway_ride <- factor(grid$average_length_subway_ride, 
                                          ordered = TRUE,
                                          levels = levels(regression_rider_data$average_length_subway_ride))
grid$frequency_of_delays <- factor(grid$frequency_of_delays, 
                                   ordered = TRUE,
                                   levels = levels(regression_rider_data$frequency_of_delays))
grid$approximate_delay_duration <- factor(grid$approximate_delay_duration, 
                                         ordered = TRUE,
                                         levels = levels(regression_rider_data$approximate_delay_duration))
grid$frequency_of_rerouting <- factor(grid$frequency_of_rerouting, 
                                      ordered = TRUE,
                                      levels = levels(regression_rider_data$frequency_of_rerouting))

#Get probabilities 
pred_probs <- predict(model, newdata = grid, type = "prob")
pred_df <- cbind(grid, pred_probs)

pred_long <- pred_df %>%
  pivot_longer(
    cols = starts_with("fit"), 
    names_to = "Satisfaction", 
    values_to = "Probability"
  )

#Recode names
pred_long <- pred_long %>%
  mutate(Satisfaction = recode(
    Satisfaction,
    "fit.1" = "Highly Unsatisfied",
    "fit.2" = "Not satisfied",
    "fit.3" = "Neutral",
    "fit.4" = "Somewhat satisfied",
    "fit.5" = "Very satisfied"
  ))


#Produce plot
plot_ly(pred_long, 
    x = ~average_length_subway_ride, 
    y = ~borough, 
    z = ~Probability, 
    type = "heatmap", 
    colorscale = "Viridis", 
    text = ~paste("Satisfaction:", Satisfaction, "<br>Probability:", round(Probability, 3)),
    hoverinfo = "text") %>%
       layout(
          title = "Predicted Probability Heatmap",
          xaxis = list(
            title = "Average Length of Subway Ride",
            tickvals = 0:5,                 
            ticktext = c("< 20 min", "20 - 40 min", "40 - 60 min", "60 - 90 min", "90 min - 2 hours")  #Putting labels back 
          ),
          yaxis = list(
            title = "Borough",
            tickvals = 0:5,                           # positions along y-axis
            ticktext = c("Manhattan", "Other Boroughs in NYC", "Outside NY")  # custom labels
          )
        )
```

