---
output: html_document 
---

By borough / zip code (like the TA’s example) 
Primary use of subway 
Average subway ride length 
Most used line 

Perception of affordability 
Opinion on increased fair 
Satisfaction 

Converting zip codes into boroughs. I used https://www.nyc.gov/assets/planning/download/office/data-maps/nyc-population/census2000/sf1p11.xls to get majority of the borough's zip codes. Any remaining ones, I looked up on Google. There were three zip codes that I could not find a match for, and were likely miswritten, so I removed those data rows. 

```{r}
library(tidyverse)
library(plotly)
library(dplyr)


zip_code_data = 
  read_csv("data/nyc_zipcodes.csv") %>%
  select(borough, zipcode)

#Make sure character type
rider_data <- rider_data %>%
  mutate(zip_code = as.character(zip_code))
zip_code_data <- zip_code_data %>%
  mutate(zipcode = as.character(zipcode))

#Join
rider_data_full <- rider_data %>%
  left_join(zip_code_data, by = c("zip_code" = "zipcode"))

#If there's no match, we want to keep the zip_code in the borough column, so we use coalesce to handle this
rider_data_full <- rider_data_full %>%
  mutate(borough = coalesce(borough, as.character(zip_code)))

#manually go through remaining zip codes - removing 11135 11290 12105 as they are not real zip codes
rider_data_full <- rider_data_full %>%
  filter(zip_code != 12105) %>%
  filter(zip_code != 11290) %>%
  filter(zip_code != 11135) 

```


Now I want to create visualizations for each variable comparison. This will demonstrate exploratory analysis and demographics. 
```{r}
#Location distribution 
borough_df <- rider_data_full %>% 
  count(borough)

plot_ly(
  borough_df,
  x = ~borough,
  y = ~n,
  type = "bar",
  hovertemplate = "Borough: %{x}<br>Count: %{y}<extra></extra>"
) %>%
  layout(title = "Survey Respondents by Borough",
         xaxis = list(title = "Borough"),
         yaxis = list(title = "Count"))


#Primary use of subway 
primary_df <- rider_data_full %>%
  count(borough, primary_use_of_subway) %>%
  group_by(borough) %>%
  mutate(prop = n / sum(n))

plot_ly(
  primary_df,
  x = ~borough,
  y = ~prop,
  color = ~primary_use_of_subway,
  type = "bar",
  hovertemplate = 
    "Borough: %{x}<br>Use: %{customdata}<br>Percent: %{y:.1%}<extra></extra>",
  customdata = ~primary_use_of_subway
) %>%
  layout(
    barmode = "stack",
    title = "Primary Subway Use by Borough (Proportions)",
    yaxis = list(tickformat = ".0%", title = "Proportion")
  )


#Average subway ride length 
ride_df <- rider_data_full %>% count(average_length_subway_ride)

plot_ly(
  ride_df,
  x = ~average_length_subway_ride,
  y = ~n,
  type = "bar",
  hovertemplate = "%{x}<br>Count: %{y}<extra></extra>"
) %>%
  layout(title = "Ride Length Distribution",
         xaxis = list(title = "Ride Length Category"),
         yaxis = list(title = "Count"))


#Most used line 
line_df <- rider_data_full %>% 
  count(subway_line_used_most_often)

plot_ly() %>%
  add_segments(
    data = line_df,
    x = 0,
    xend = ~n,
    y = ~reorder(subway_line_used_most_often, n),
    yend = ~reorder(subway_line_used_most_often, n),
    hoverinfo = "none"
  ) %>%
  add_markers(
    data = line_df,
    x = ~n,
    y = ~reorder(subway_line_used_most_often, n),
    hovertemplate = "Line: %{y}<br>Count: %{x}<extra></extra>"
  ) %>%
  layout(
    title = "Most Used Subway Lines",
    xaxis = list(title = "Count"),
    yaxis = list(title = "Line")
  )


#Perception of affordability 
afford_df <- rider_data_full %>%
  count(borough, is_subway_affordable) %>%
  group_by(borough) %>%
  mutate(prop = n / sum(n))

plot_ly(
  afford_df,
  x = ~borough,
  y = ~prop,
  color = ~is_subway_affordable,
  type = "bar",
  hovertemplate = 
    "Borough: %{x}<br>Affordability: %{customdata}<br>Percent: %{y:.1%}<extra></extra>",
  customdata = ~is_subway_affordable
) %>%
  layout(
    barmode = "stack",
    title = "Affordability Perception by Borough",
    yaxis = list(tickformat = ".0%", title = "Proportion")
  )


#Opinion on increased fair 
satis_df <- rider_data_full %>%
  count(borough, opinion_on_increased_fares) %>%
  group_by(borough) %>%
  mutate(
    prop = n / sum(n),
    percent_label = paste0(round(prop * 100, 1), "%")
  ) %>%
  ungroup()

library(plotly)

boroughs <- unique(satis_df$borough)
plot <- plot_ly()

for (i in seq_along(boroughs)) {
  
  b <- boroughs[i]
  df_b <- filter(satis_df, borough == b)
  
  plot <- add_trace(
    plot,
    data = df_b,
    labels = ~opinion_on_increased_fares,
    values = ~prop,
    type = "pie",
    hole = 0.2 + (i - 1) * 0.15,
    textinfo = "none",
    hovertemplate = paste(
      "<b>Borough:</b> ", b, "<br>",
      "<b>Opinion:</b> %{label}<br>",
      "<b>Percent:</b> %{customdata}<extra></extra>"
    ),
    customdata = ~percent_label,
    showlegend = (i == 1)
  )
}

plot <- layout(
  plot,
  title = "Opinion on Increased Fares (Multi-Ring Donut)",
  margin = list(l=20, r=20, t=60, b=20)
)

plot


#Satisfaction 
satis_overall <- rider_data_full %>%
  count(overall_satisfaction) %>%
  mutate(prop = n / sum(n),
         percent = paste0(round(prop*100,1), "%"))

plot_ly(
  satis_overall,
  labels = ~overall_satisfaction,
  values = ~prop,
  type = "pie",
  hole = .5,
  textinfo = "label+percent",
  hovertemplate = "%{label}<br>%{percent}<extra></extra>"
) %>%
  layout(title = "Overall Satisfaction Levels")
```


```{r}
#interactive plotly for satisfaction 

satis_df <- rider_data_full %>%
  count(borough, overall_satisfaction) %>%
  group_by(borough) %>%
  mutate(
    prop = n / sum(n),
    percent_label = paste0(round(prop*100, 1), "%")
  )
library(plotly)
plots <- lapply(unique(satis_df$borough), function(b) {
  df_b <- filter(satis_df, borough == b)
  
  plot_ly(
    df_b,
    labels = ~overall_satisfaction,
    values = ~prop,
    type = "pie",
    hole = 0.5,
    textinfo = "none",
    hovertemplate = paste(
      "<b>Borough:</b> ", b, "<br>",
      "<b>Category:</b> %{label}<br>",
      "<b>Percent:</b> %{customdata}<extra></extra>"
    ),
    customdata = ~percent_label
  ) %>%
    layout(title = b, showlegend = FALSE)
})

subplot(plots, nrows = 2, margin = 0.02) %>%
  layout(title = "MTA Satisfaction Levels by Borough")



```


```{r}
satis_df <- rider_data_full %>%
  count(borough, overall_satisfaction) %>%
  group_by(borough) %>%
  mutate(
    prop = n / sum(n),
    percent_label = paste0(round(prop * 100, 1), "%")
  ) %>%
  ungroup()


library(plotly)

boroughs <- unique(satis_df$borough)

p <- plot_ly()

for (i in seq_along(boroughs)) {
  
  b <- boroughs[i]
  df_b <- filter(satis_df, borough == b)
  
  # Each borough gets its own ring
  p <- add_trace(
    p,
    data = df_b,
    labels = ~overall_satisfaction,
    values = ~prop,
    type = "pie",
    hole = 0.2 + (i - 1) * 0.15,      # inner hole size per ring
    sort = FALSE,
    direction = "clockwise",
    textinfo = "none",
    hovertemplate = paste(
      "<b>Borough:</b> ", b, "<br>",
      "<b>Category:</b> %{label}<br>",
      "<b>Percent:</b> %{customdata}<extra></extra>"
    ),
    customdata = ~percent_label,
    showlegend = (i == 1)             # legend only once
  )
}

p <- layout(
  p,
  title = "MTA Satisfaction Across Boroughs (Multi-Ring Donut)",
  margin = list(l = 20, r = 20, t = 50, b = 20)
)

p

```

