---
output: github_document
---

By borough / zip code (like the TA’s example) 
Primary use of subway 
Average subway ride length 
Most used line 

Perception of affordability 
Opinion on increased fair 
Satisfaction 

Converting zip codes into boroughs. I used https://www.nyc.gov/assets/planning/download/office/data-maps/nyc-population/census2000/sf1p11.xls to get majority of the borough's zip codes. Any remaining ones, I looked up on Google. There were three zip codes that I could not find a match for, and were likely miswritten, so I removed those data rows. 

```{r}
zip_code_data = 
  read_csv("data/nyc_zipcodes.csv") %>%
  select(borough, zipcode)

#Make sure character type
rider_data <- rider_data %>%
  mutate(zip_code = as.character(zip_code))
zip_code_data <- zip_code_data %>%
  mutate(zipcode = as.character(zipcode))

#Join
rider_data_full <- rider_data %>%
  left_join(zip_code_data, by = c("zip_code" = "zipcode"))

#If there's no match, we want to keep the zip_code in the borough column, so we use coalesce to handle this
rider_data_full <- rider_data_full %>%
  mutate(borough = coalesce(borough, as.character(zip_code)))

#manually go through remaining zip codes - removing 11135 11290 12105 as they are not real zip codes
rider_data_full <- rider_data_full %>%
  filter(zip_code != 12105) %>%
  filter(zip_code != 11290) %>%
  filter(zip_code != 11135) 
```


Now I want to create visualizations for each variable comparison. This will demonstrate exploratory analysis and demographics. 
```{r}
#Primary use of subway 

#Average subway ride length 


#Most used line 

#Perception of affordability 

#Opinion on increased fair 

#Satisfaction 

```

