---
title: "Subway Line Analysis"
output: html_document
---


```{r, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(lubridate)
library(plotly)
library(viridis)
library(readxl)
library(treemapify)
library(dplyr)
library(gt)
```
<br>
The Subway Line Analysis explores how rider experiences, delay patterns, and service perceptions differ across New York City’s subway lines. Our goal is to identify which lines offer smoother commutes, which struggle with chronic delays, and how these differences shape overall rider satisfaction.

Across multiple visualizations, we examine how frequently riders on each line report delays, how often those delays exceed 20 minutes, and how complaint patterns vary by route. The composite score table summarizes overall performance across four key service indicators, offering a system-wide ranking of line families. Together, these interactive graphics reveal both consistent system-wide trends, such as unusually high everyday delays on lines like the 7 or the severe long-delay burden on the Z train. 



```{r, include=FALSE}
#Clean data
subway_rider_data = 
  read_csv("data/2019_subway_rider_data.csv") |> 
  janitor::clean_names()


subway_drops =
  subway_rider_data |> 
  select(-survey_stop_borough, -survey_stop_location) |> 
  drop_na() |> 
  mutate(survey_id = row_number()) |> 
  relocate(survey_id, .before = 1)
  
rider_data =
  subway_drops |>
  mutate(
    use_of_subway_frequency = case_when(
      use_of_subway_frequency == "Everyday / Almost everyday" ~ "Every day",
      use_of_subway_frequency == "More than once a week, varies" ~ "Several times a week",
      use_of_subway_frequency %in% c("Weekdays, Mon - Fri", "~5 days per week") ~ "5 days per week",
      use_of_subway_frequency == "Once a week" ~ "Once a week",
      use_of_subway_frequency == "Once a month" ~ "Once a month",
      use_of_subway_frequency == "Almost never" ~ "Almost never",
      use_of_subway_frequency == "Never" ~ "Never",
      TRUE ~ NA_character_
    ),
    use_of_subway_frequency = 
      factor(
        use_of_subway_frequency,
        levels = c("Never", 
                   "Almost never", 
                   "Once a month", 
                   "Once a week",
                   "Several times a week", 
                   "5 days per week", 
                   "Every day"),
        ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(average_length_subway_ride = 
           factor(average_length_subway_ride,
                  levels = c("< 20 min", 
                             "20 - 40 min", 
                             "40 - 60 min",
                             "60 - 90 min", 
                             "90 min - 2 hours"),
                  ordered = TRUE))


rider_data =
  subway_drops |>
  mutate(
    overall_satisfaction = 
      recode(
        overall_satisfaction,
        "Not satisfied" = "Unsatisfied"),
    overall_satisfaction = 
      factor(
        overall_satisfaction,
        levels = c("Highly unsatisfied",
                 "Unsatisfied",
                 "Neutral",
                 "Somewhat satisfied",
                 "Very satisfied"),
      ordered = TRUE)
    )


rider_data =
  subway_drops |> 
  mutate(
    frequency_of_delays = factor(
      frequency_of_delays,
      levels = c("Everyday",
                 "A few times a week",
                 "Rarely",
                 "Almost never",
                 "Never"),
      ordered = TRUE)
    )
    
#Zip code --> borough 

zip_code_data = 
  read_csv("data/nyc_zipcodes.csv") %>%
  select(borough, zipcode)

#Make sure character type
rider_data <- rider_data %>%
  mutate(zip_code = as.character(zip_code))
zip_code_data <- zip_code_data %>%
  mutate(zipcode = as.character(zipcode))

#Join
rider_data_full <- rider_data %>%
  left_join(zip_code_data, by = c("zip_code" = "zipcode"))

#If there's no match, we want to keep the zip_code in the borough column, so we use coalesce to handle this
rider_data_full <- rider_data_full %>%
  mutate(borough = coalesce(borough, as.character(zip_code)))

#manually go through remaining zip codes - removing 11135 11290 12105 as they are not real zip codes & also removing one observation that didn't have zip code 
rider_data_full <- rider_data_full %>%
  filter(survey_id != 2410) %>% 
  filter(survey_id != 944) %>% 
  filter(survey_id != 241) %>% 
  filter(survey_id != 919) 

# Creating a complaint_data_time dataset (I added this)
complaint_data_time <- rider_data_full |> 
  mutate(
    submitted_at = mdy_hm(submitted_at),
    hour = hour(submitted_at),
    time_of_day = case_when(
      hour >= 6 & hour < 12 ~ "Morning",  #6 am to 11:59 am
      hour >= 12 & hour < 17 ~ "Afternoon", #12 pm to 4:59 pm
      hour >= 17 & hour < 21 ~ "Evening",  #5pm to 8:59 pm
      TRUE ~ "Night" # 9pm to 5:59 am 
    ),
    time_of_day = factor(time_of_day, 
                        levels = c("Morning", "Afternoon", "Evening", "Night"))
  )


```


```{r, include=FALSE}
# Subway Line Directory Table Code:

subway_directory_display <- tribble(
  ~`Line Family`, ~Lines, ~Color,
  "Red Line", "1, 2, 3", "#EE352E",
  "Green Line", "4, 5, 6", "#00933C",
  "Purple Line", "7", "#B933AD",
  "Blue Line", "A, C, E", "#0039A6",
  "Orange Line", "B, D, F, M", "#FF6319",
  "Yellow Line", "N, Q, R, W", "#FCCC0A",
  "Light Green Line", "G", "#6CBE45",
  "Brown Line", "J, Z", "#996633",
  "Gray Line", "L", "#A7A9AC",
  "Shuttle", "S", "#808183"
)

subway_directory_display %>%
  gt() %>%
  tab_header(title = "NYC Subway Line Directory") %>%
  text_transform(
    locations = cells_body(columns = Color),
    fn = function(x) {
      # x = column vector of hex colors
      mapply(function(col) {
        html(
          paste0(
            "<div style='background-color:", col,
            "; width: 60px; height: 20px; border-radius: 4px'></div>"
          )
        )
      }, x)
    }
  ) %>%
  cols_width(
    `Line Family` ~ px(180),
    Lines ~ px(180),
    Color ~ px(100)
  ) %>%
  opt_table_outline()
```


```{r, include=FALSE}
# Join-ready Subway Line Directory + Join to cleaned dataset
subway_directory <- tribble(
~subway_line, ~group_color, ~hex_color,
"1","Red Line","#EE352E",
"2","Red Line","#EE352E",
"3","Red Line","#EE352E",
"4","Green Line","#00933C",
"5","Green Line","#00933C",
"6","Green Line","#00933C",
"7","Purple Line","#B933AD",
"A","Blue Line","#0039A6",
"C","Blue Line","#0039A6",
"E","Blue Line","#0039A6",
"B","Orange Line","#FF6319",
"D","Orange Line","#FF6319",
"F","Orange Line","#FF6319",
"M","Orange Line","#FF6319",
"N","Yellow Line","#FCCC0A",
"Q","Yellow Line","#FCCC0A",
"R","Yellow Line","#FCCC0A",
"W","Yellow Line","#FCCC0A",
"G","Light Green Line","#6CBE45",
"J","Brown Line","#996633",
"Z","Brown Line","#996633",
"L","Gray Line","#A7A9AC",
"S","Shuttle","#808183"
)
```


```{r, include=FALSE}
# Categories defining "long delays"
long_categories <-  c(
  "20 - 40 min", "40 - 60 min", "60 - 90 min",
  "90 min - 2 hours", "> 60 min"
)

# Ordinal score for frequency of delays
delay_freq_scores <- c(
  "Never" = 1,
  "Almost never" = 2,
  "Rarely" = 3,
  "A few times a week" = 4,
  "Everyday" = 5
)

# Numeric midpoints for average delay length
delay_length_scores <- c(
  "< 20 min"        = 10,     # midpoint choice
  "20 - 40 min"     = 30,
  "40 - 60 min"     = 50,
  "60 - 90 min"     = 75,
  "90 min - 2 hours" = 105
)

# Satisfaction ordinal scale
satisfaction_scores <- c(
  "Highly unsatisfied" = 1,
  "Unsatisfied"        = 2,
  "Neutral"            = 3,
  "Somewhat satisfied" = 4,
  "Very satisfied"     = 5
)
```

```{r, include=FALSE}
scored_data <- rider_data_full |>
  mutate(
    subway_line = subway_line_used_most_often |>
      str_replace_all("[^A-Za-z0-9]", "/") |>
      str_split("/")
  ) |>
  unnest(subway_line) |>
  mutate(
    subway_line = str_trim(subway_line),
    subway_line = str_to_upper(subway_line)
  ) |>
  filter(subway_line %in% subway_directory$subway_line) |>
  left_join(subway_directory, by = "subway_line") |>
  mutate(
    long_delay       = average_length_subway_ride %in% long_categories,
    delay_freq_num   = delay_freq_scores[frequency_of_delays],
    delay_length_num = delay_length_scores[average_length_subway_ride],   # ← FIX
    satisfaction_num = satisfaction_scores[overall_satisfaction]
  )

family_scores_raw <- scored_data |>
  group_by(group_color) |>
  summarise(
    pct_long_delay   = mean(long_delay, na.rm = TRUE),
    avg_delay_freq   = mean(delay_freq_num, na.rm = TRUE),
    avg_delay_length = mean(delay_length_num, na.rm = TRUE),
    avg_satisfaction = mean(satisfaction_num, na.rm = TRUE)
  ) |>
  ungroup()

family_scores_norm <- family_scores_raw |>
  mutate(
    z_pct_long_delay   = scale(pct_long_delay)[,1]    * -1,
    z_delay_freq       = scale(avg_delay_freq)[,1]    * -1,
    z_delay_length     = scale(avg_delay_length)[,1]  * -1,
    z_satisfaction     = scale(avg_satisfaction)[,1]
  )

family_composite <- family_scores_raw |>
  mutate(
    z_pct_long_delay = scale(-pct_long_delay)[,1],
    z_delay_freq     = scale(-avg_delay_freq)[,1],
    z_delay_length   = scale(-avg_delay_length)[,1],
    z_satisfaction   = scale(avg_satisfaction)[,1]
  ) |>

  # Replace z-score NAs with 0 (neutral)
  mutate(
    across(starts_with("z_"), ~replace_na(.x, 0))
  ) |>

  mutate(
    composite_score = z_pct_long_delay +
                      z_delay_freq +
                      z_delay_length +
                      z_satisfaction
  ) |>
  arrange(desc(composite_score)) |>
  mutate(rank = row_number())
```

```{r, include=FALSE}
# Step 2: Remove extra columns & make all scores positive
family_composite_clean <- family_composite |>
  mutate(
    composite_score = composite_score - min(composite_score, na.rm = TRUE)
  ) |>
  arrange(desc(composite_score)) |>
  mutate(rank = row_number()) |>
  select(group_color, composite_score, rank)

subway_directory_display_final <- subway_directory_display |>
  left_join(family_composite_clean, by = c("Line Family" = "group_color"))
```


```{r}
subway_directory_display_final =
  subway_directory_display_final %>%
  arrange(rank) %>%
  mutate(
    rank = dplyr::case_when(
      rank == 1 ~ "🥇 1",
      rank == 2 ~ "🥈 2",
      rank == 3 ~ "🥉 3",
      TRUE      ~ as.character(rank)
    )
  ) %>%
  gt() %>%
  tab_header(title = "NYC Subway Line Directory with Composite Score") %>%
  cols_move(
    columns = c(composite_score, rank),
    after = Color
  ) %>%
  cols_label(
    composite_score = "Score",
    rank            = "Rank"
  ) %>%
  fmt_number(
    columns = composite_score,
    decimals = 2
  ) %>%
  text_transform(
    locations = cells_body(columns = Color),
    fn = function(x) {
      mapply(function(col) {
        html(
          paste0(
            "<div style='background-color:", col,
            "; width: 60px; height: 20px; border-radius: 4px'></div>"
          )
        )
      }, x)
    }
  ) %>%
  cols_width(
    `Line Family`   ~ px(120),
    Lines           ~ px(100),
    Color           ~ px(70),
    composite_score ~ px(80),
    rank            ~ px(70)
  ) %>%
  cols_align(
    align = "center",
    columns = rank
  ) %>%
  opt_table_outline()

subway_directory_display_final

```

<br><br>

<h3>How often are these trains delayed?</h3>

```{r, fig.cap="This heat map highlights system-wide patterns, showing consistently high weekly delay reports and notable everyday delays on lines like the G and the Z. Hover details reveal the precise counts and proportions underlying each cell."}
# Distribution of delay frequency -- heat map

keep_delays <- c("Rarely", "A few times a week", "Everyday")

heat_data <- scored_data |>
  filter(!is.na(frequency_of_delays)) |>
  filter(frequency_of_delays %in% keep_delays) |>
  count(subway_line, group_color, frequency_of_delays, name = "n") |>
  complete(
    subway_line = unique(subway_directory$subway_line),
    frequency_of_delays = keep_delays,
    fill = list(n = 0)
  ) |>
  group_by(subway_line) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  left_join(subway_directory, by = "subway_line") |>
  mutate(
    subway_line = factor(subway_line, levels = unique(subway_line)),
    frequency_of_delays = factor(frequency_of_delays, levels = keep_delays)
  )

heat_data_plot <- plot_ly(
  data = heat_data,
  x = ~frequency_of_delays,
  y = ~subway_line,
  z = ~prop,
  type = "heatmap",
  colors = viridisLite::viridis(100),
  customdata = ~n,
  hovertemplate = paste(
    "Line: %{y}<br>",
    "Delay Frequency: %{x}<br>",
    "Proportion: %{z:.1%}<br>",
    "Count: %{customdata}<br>",
    "<extra></extra>"
  )
) |>
  layout(
    title = "Delay Frequency Distribution by Subway Line",
    xaxis = list(title = "Delay Frequency"),
    yaxis = list(title = "Subway Line"),
    margin = list(l = 110, r = 20, b = 80, t = 80)
  )

heat_data_plot
```


<br><br>
<h3>Are riders experiencing delays?</h3>

```{r}

# To establish proportion of Riders with Long Delays (≥20 min): Define long delay buckets.

# Proportion of Riders with Long Delays (≥20 min)
# Interactive plotly bar chart

long_categories <- c(
  "20 - 40 min", "40 - 60 min", "60 - 90 min",
  "90 min - 2 hours", "> 60 min"
)

line_order <- subway_directory |>
  arrange(group_color, subway_line) |>
  pull(subway_line)

long_delay_by_line <- scored_data |>
  filter(!is.na(average_length_subway_ride)) |>
  mutate(long_delay = average_length_subway_ride %in% long_categories) |>
  count(subway_line, group_color, hex_color, long_delay, name = "n") |>
  group_by(subway_line, group_color, hex_color) |>
  mutate(prop = n / sum(n)) |>
  filter(long_delay) |>
  ungroup() |>
  mutate(
    subway_line = factor(subway_line, levels = line_order)
  )

plot_ly(
  data = long_delay_by_line,
  x = ~subway_line,
  y = ~prop,
  type = "bar",
  marker = list(color = ~hex_color),
  customdata = ~group_color,
  hovertemplate = paste(
    "<b>Subway Line:</b> %{x}<br>",
    "<b>Long Delay Proportion:</b> %{y:.2f}<br>",
    "<b>Line Group:</b> %{customdata}<br>",
    "<extra></extra>"
  )
) |>
  layout(
    title = "Proportion of Riders Reporting Long Delays (≥20 min)",
    xaxis = list(title = "Subway Line", categoryarray = line_order),
    yaxis = list(title = "Proportion", tickformat = ".0%"),
    bargap = 0.20
  )

```

This bar chart shows the share of riders on each line who report experiencing long delays of 20 minutes or longer. While most lines cluster between 60-80%, and the Z line lacked enough data to note high delay proportions. Interactive hover labels display each line’s exact proportion and color group for more detailed comparison.

<br><br>

<h3>Which trains are the worst with delays?</h3>

```{r, include=FALSE}
delay_counts =
  rider_data |>
  filter(frequency_of_delays %in% c("Everyday", "A few times a week")) |>
  count(subway_line_used_most_often, sort = TRUE) |>
  left_join(
    subway_directory,
    by = c("subway_line_used_most_often" = "subway_line")
  )
```


```{r, fig.cap= "The 1 Train seems to have a signficant amount of delays compare to the other trains. It has double the amount of complaints compared to the second train with the most complaints, the R train. Is there a reason why these trains seem to have the most delay issues? I guess we will have to do a future survey to find out."}
delay_counts_plot =
  plot_ly(
    delay_counts,
    labels = ~subway_line_used_most_often,
    values = ~n,
    type   = "pie",
    marker = list(
      colors = delay_counts |> dplyr::pull(hex_color)
    ),
    text = ~paste0(subway_line_used_most_often, " Train"),
    textinfo = "percent",     
    textposition = "inside",
    hovertext = ~paste0(
      "Line: ", subway_line_used_most_often,
      " (", group_color, ")",
      "<br>Count: ", n
    ),
    hoverinfo = "text",
    showlegend = FALSE            
  ) |>
  layout(
    title = "Subway Lines with the Most Delays",
    margin = list(l = 40, r = 40, b = 40, t = 80)
  )

delay_counts_plot




```

<br><br>

<h3>What are the subway complaints reported?</h3>


```{r}
# I added this ( adding the complaint variable)
rider_data_full <- rider_data_full |>
  mutate(
    complaint = str_split(top_three_complaints, ",\\s*")
  ) |>
  tidyr::unnest(complaint)
```


```{r}
# Creating a complaint_data_time dataset (I added this)
complaint_data_time <- rider_data_full |> 
  mutate(
    submitted_at = mdy_hm(submitted_at),
    hour = hour(submitted_at),
    time_of_day = case_when(
      hour >= 6 & hour < 12 ~ "Morning",  #6 am to 11:59 am
      hour >= 12 & hour < 17 ~ "Afternoon", #12 pm to 4:59 pm
      hour >= 17 & hour < 21 ~ "Evening",  #5pm to 8:59 pm
      TRUE ~ "Night" # 9pm to 5:59 am 
    ),
    time_of_day = factor(time_of_day, 
                        levels = c("Morning", "Afternoon", "Evening", "Night"))
  )
```



```{r, fig.width=9, fig.height=6, fig.cap= " This chart provides a detailed breakdown of subway complaints by both the specific transit line and the time of day (morning, afternoon, evening, night).The stacked bar graphs reveal how the distribution of primary complaints (such as delays, overcrowding, and crime) shifts significantly throughout the day, highlighting the distinct challenges during peak versus off-peak hours.The visualization allows for clear comparison across the system, showing which issues are most pressing for each line at different times." } 

complaint_summary <- complaint_data_time |> 
  count(subway_line_used_most_often, time_of_day, complaint) |> 
  group_by(subway_line_used_most_often, time_of_day) |> 
  mutate(percent = n / sum(n) * 100) |> 
  ungroup()


library(viridis)

all_complaints <- sort(unique(complaint_summary$complaint))
viridis_colors <- viridis(length(all_complaints))
names(viridis_colors) <- all_complaints


times <- unique(complaint_summary$time_of_day)

plot <- plot_ly()
i <- 0

for (t in times) {
  df_t <- complaint_summary |> filter(time_of_day == t)
  
  for (comp in unique(df_t$complaint)) {
    df_tc <- df_t |> filter(complaint == comp)
    
    plot <- plot |> add_trace(
      data = df_tc,
      x = ~percent,
      y = ~subway_line_used_most_often,
      type = "bar",
      orientation = "h",
      name = comp,
      hovertemplate = paste(
        "Time: ", t,
        "<br>Line: %{y}",
        "<br>Complaint: ", comp,
        "<br>Count: %{customdata}",
        "<br>Percent: %{x:.1f}%",
        "<extra></extra>"
      ),
      customdata = ~n,
      marker = list(color = viridis_colors[comp]),
      visible = ifelse(t == times[1], TRUE, FALSE),
      legendgroup = comp
    )
    
    i <- i + 1
  }
}

# Build dropdown buttons
buttons <- list()
for (k in seq_along(times)) {
  vis <- rep(FALSE, length(unique(complaint_summary$complaint)) * length(times))
  
  start <- (k - 1) * length(unique(complaint_summary$complaint)) + 1
  end <- start + length(unique(complaint_summary$complaint)) - 1
  
  vis[start:end] <- TRUE
  
  buttons[[k]] <- list(
    method = "restyle",
    args = list("visible", vis),
    label = times[k]
  )
}

# Final layout with dropdown moved
plot |> layout(
  title = "Complaint Percent by Subway Line and Time of Day",
  xaxis = list(title = "Percent of Complaints"),
  yaxis = list(title = "Subway Line"),
  barmode = "stack",
  # Move legend if needed (optional)
  legend = list(
    x = 1.1,
    y = 0.5
  ),
  updatemenus = list(
    list(
      y = 0.98,        # Move to top right, below title
      x = 0.98,        # Right side
      buttons = buttons,
      type = "dropdown",
      xanchor = "left",  # Anchor to left side
      yanchor = "top"     # Anchor to top
    )
  )
)

```



























